Black-swan events in animal populations

### 2016-11-11 text/ms_pnas.tex
Copyediting

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012}, including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction.
Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.
The prevalence of population die-offs rather than unexpected increases
is likely due to biology:
there is no limit to how fast a population can decline due to a catastrophe,
but population increases are constrained
by the maximum biological rate of increase
for a given set of demographic traits
(e.g. population birth rate, mortality, age at maturity).

There are a number of caveats when considering the generality of our results.
First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Second, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors (Supporting Information).
Third, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
Finally, our models assume that
the fitted parameters (e.g. productivity and density dependence)
do not vary over time.
Allowing for time varying parameters
could be an important development for
understanding the dynamics of longer time series.

>>>>

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a $t$ distribution with heavy
tails \cite{gelman2014}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \cite{schindler2010}, or an intrinsic change in population variability
through time \cite{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \cite{nunez2012} including extreme climate
events \cite{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black-swan events, or sharp changes in human
pressures such as fishing or habitat destruction.
Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \cite{kirby2009} or a rare alignment of
non-synergistic forces \cite{denny2009}.
The prevalence of population die-offs rather than unexpected increases
is likely due to biology:
there is no limit to how fast a population can decline due to a catastrophe,
but population increases are constrained
by the maximum biological rate of increase
for a given set of demographic traits
(e.g. population birth rate, mortality, age at maturity).

There are a number of caveats when considering the generality of our results.
First, the Global Population Dynamics Database \cite[GPDD,][]{gpdd2010} we used
represents a taxonomically and geographically biased sample of
populations---the longer time series we focused on are dominated by commercially
and recreationally important species and a disproportionate number of
populations are located in the United Kingdom. Although we would expect to find
qualitatively similar evidence for black swans in other large taxonomic or
geographic samples of populations, the common forces driving those black swans
would likely differ. Second, some apparent black-swan events could be
recording mistakes, although conversely, some extreme observations may have been
discarded or altered if they were erroneously suspected of being recording
mistakes. Indeed, we discarded three of the populations that our method
initially identified as heavy tailed because they turned out to be data-entry
errors (Supporting Information).
Third, the temporal scales of observation and population dynamics
vary considerably across populations in the GPDD and these likely influence the
detection of heavy tails. For example, if we make frequent observations
relative to generation time (e.g.~for many large-bodied mammals) we will
average across generations and perhaps miss black-swan events. Conversely, if
we census populations infrequently relative to generation time (e.g.~many
insects in the GPDD) the recorded data may average across extreme and
less-extreme events and also dampen black-swan dynamics.
Finally, our models assume that
the fitted parameters (e.g. productivity and density dependence)
do not vary over time.
Allowing for time-varying parameters
could be an important development for
understanding the dynamics of longer time series.
